---
title: "Linking Expat from another package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linking Expat from another package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Who this is for

Read this if your package has C code that is written against **Expat itself** ---
it calls `XML_ParserCreate()`, `XML_GetBuffer()`, `XML_StopParser()` --- and you
want it to build without a system `libexpat`. Typically the code is vendored and
cannot be rewritten: `xlsxio`, which `zuxlsx` vendors, is that case.

If instead you are writing new C code that you control, **stop here and read
`vignette("streaming")`** first. The registered function table described there
is the better path for new code: it needs no `configure`, no linker flags, and
picks up a zuxml fix when zuxml is upgraded. This vignette is for the case where
that is not an option.

## The two modes

|                        | table (`zuxml_api`)              | archive (`libzuxml.a`)          |
|------------------------|----------------------------------|---------------------------------|
| `DESCRIPTION`          | `Imports:` **and** `LinkingTo:`  | `LinkingTo:` only               |
| `NAMESPACE`            | needs an `importFrom()`          | nothing                         |
| Header                 | `zuxml.h`, no Expat type in it   | `expat.h`, Expat's own API      |
| API style              | callbacks through a table        | everything Expat offers         |
| Symbols resolved       | at run time, `R_GetCCallable()`  | at link time, into your `.so`   |
| zuxml at run time      | installed **and** loadable       | not needed at all               |
| A zuxml fix reaches you| when zuxml is upgraded           | when *you* are reinstalled      |

The archive exists because the table cannot express a *pull*-style reader. A
reader that returns a row to its caller and resumes where it left off is built
on `XML_StopParser()`/`XML_ResumeParser()`, and there is no equivalent in
`zuxml.h`.

## What an installed zuxml gives you

```{r}
# system.file() returns "" outside an installed layout, and dir("") would then
# list the working directory rather than nothing.
installed <- function(what) {
  path <- system.file(what, package = "zuxml")
  if (nzchar(path)) dir(path) else "(not an installed layout)"
}

installed("include")
installed("lib")
```

`LinkingTo: zuxml` puts that `include` directory on your compiler's path
(R appends it to `CLINK_CPPFLAGS`), so `#include <expat.h>` resolves with no
further work.

The headers are not kept under `inst/include/` in zuxml's sources. They are
copied out of the vendored tree at install time by `src/install.libs.R`, so the
header you compile against is by construction the one the objects in the
archive were compiled from.

## There is no `LinkingTo` for a library

`LinkingTo` handles headers and nothing else. For the archive you have to find
`system.file("lib", package = "zuxml")` yourself, and the only portable place to
do that is a `configure` script that writes `src/Makevars`.

The two shortcuts are both worse. `$(shell ...)` in `Makevars` forces
`SystemRequirements: GNU make`. An `Imports: zuxml` entry gets you nothing:
`Imports` is about loading a namespace at run time, which a statically linked
consumer never does.

### `configure`

```sh
#!/bin/sh
set -eu

: "${R_HOME:?configure must be run by R CMD INSTALL, which sets R_HOME}"
RSCRIPT="${R_HOME}/bin/Rscript"

ZUXML_LIB=$("${RSCRIPT}" --vanilla -e "cat(system.file('lib', package = 'zuxml'))")

if [ -z "${ZUXML_LIB}" ] || [ ! -f "${ZUXML_LIB}/libzuxml.a" ]; then
  echo "configure: zuxml/lib/libzuxml.a was not found." >&2
  exit 1
fi

sed -e "s|@ZUXML_LIB@|${ZUXML_LIB}|g" src/Makevars.in > src/Makevars
```

Check for the **file**, not for a version. An installed zuxml from before the
archive existed reports the same version as one that carries it, so its presence
on disk is the only honest test.

`configure.win` is the same script ending in `src/Makevars.win`, and a `cleanup`
script should remove both, since they are generated:

```sh
#!/bin/sh
rm -f src/Makevars src/Makevars.win
```

### `src/Makevars.in`

```make
PKG_CPPFLAGS = -DXML_STATIC
PKG_LIBS = '@ZUXML_LIB@/libzuxml.a'
```

Two details that are easy to get wrong and fail only on one platform:

* **`-DXML_STATIC` matters on Windows.** Without it `expat_external.h` decorates
  every declaration with `__declspec(dllimport)` and the link fails.
* **Quote the path.** It comes from `system.file()`, so it lives under the R
  library, and on Windows the user library sits under the profile directory ---
  a path with a space in it reaches the linker as two nonexistent arguments. R
  quotes its own `LinkingTo` include paths for the same reason.

### `DESCRIPTION`

```
LinkingTo: zuxml
```

That is all. No `Imports: zuxml`, and nothing in `NAMESPACE`.

## Which Expat you are getting

The archive holds the Expat objects zuxml's own shared object was linked from,
and **nothing else** --- no `zux_*` wrapper, no R glue, since either would be
dead weight or a duplicate symbol inside your shared object.

It is compiled with zuxml's feature policy, which is stricter than a distro
build, and your code inherits every part of it:

* **`XML_GE 0`, and `XML_DTD` never defined.** General entities, parameter
  entities, external subsets and the external-entity machinery are compiled
  out, so XXE and entity amplification are impossible rather than disabled. The
  cost: any entity reference other than the five built-ins and numeric
  character references is a parse error.
* **Do not define `XML_GE=1` on your own command line.** `expat.h` would then
  declare `XML_SetBillionLaughsAttackProtection*()`, which the archive does not
  define, and you would get a link error. Those limiters are pointless here
  anyway --- the attack they bound is already compiled out.
* **`XML_NS 1`.** Namespace-aware parsing via `XML_ParserCreateNS()` works.
* **`XML_CONTEXT_BYTES 1024`**, so `XML_GetInputContext()` is available.
* **`XML_UNICODE` is not defined**, so `XML_Char` is plain `char` and input is
  UTF-8. `XML_LARGE_SIZE` and `XML_ATTR_INFO` are not defined either.

## Check that you actually linked it

On Linux and Windows a wrong `PKG_LIBS` fails at link time. **On macOS it does
not.** R links a package shared object with `-undefined dynamic_lookup`, so a
consumer that never linked the archive still builds, still loads, and still
parses --- against whatever Expat happens to be in the process, usually the
system one. Every behavioural test passes. The only reliable check is the
symbol table:

```sh
# Nothing may be left for the loader to resolve.
nm -u path/to/yourpkg.so | grep XML_

# And Expat has to actually be in there.
nm path/to/yourpkg.so | grep ' [TtDd] _*XML_' | wc -l
```

## A worked example

`tools/zuxmltest/` in zuxml's own sources is a complete, minimal package in
exactly this shape --- `configure`, `configure.win`, `cleanup`,
`src/Makevars.in`, and a `src/consume.c` that drives
`XML_GetBuffer`/`XML_ParseBuffer` with a suspend/resume round trip.
`tools/run-downstream-check` installs it against a freshly built zuxml and
asserts all of the above, including that it still parses after zuxml has been
removed from the library path.
