---
title: "Parsing untrusted XML"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parsing untrusted XML}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(zuxml)
```

The usual reason to parse XML is that something else sent it to you --- an HTTP
response body, a file someone uploaded, a feed you do not control. This
vignette is about what `zuxml` does with input like that, and, just as
importantly, what it refuses to do at all.

The short version: the dangerous parts of XML are not turned off by a
configuration flag you could forget to set. They are not compiled in.

## What the build actually contains

`zuxml_info()` reports the compiled-in policy rather than the current options,
so it answers "what is this binary capable of" rather than "what did you ask
for":

```{r}
zuxml_info()
```

`DTD: disabled` and `External entities: not compiled in` are the two lines that
matter. The bundled Expat is built with `XML_GE 0` and without `XML_DTD`, which
removes general entities, parameter entities and the external-entity machinery
from the object code. There is no runtime path to them because there is no
code to reach.

## XXE

The classic attack supplies a document that declares an entity pointing at a
local file, then references it, so the file's contents end up in the parsed
output:

```{r, error = TRUE}
xml_parse('<!DOCTYPE d [<!ENTITY x SYSTEM "file:///etc/passwd">]><d>&x;</d>')
```

The document is rejected at the `DOCTYPE`, before the entity declaration is
even looked at. No file is opened. The same applies to a `SYSTEM` identifier
naming a URL --- no request is made, because nothing in the package can make
one.

This is checked at two levels: `zuxml` refuses the declaration, and the parser
underneath it has no resolver compiled in. Removing either one is caught by the
package's own mutation tests, which delete each guard and require the hostile
input to stop being rejected.

## Entity expansion ("billion laughs")

The amplification attack nests entity definitions so that a small document
expands to gigabytes:

```{r, error = TRUE}
xml_parse(paste0(
  '<!DOCTYPE lol [',
  '<!ENTITY a "aaaaaaaaaa">',
  '<!ENTITY b "&a;&a;&a;&a;&a;&a;&a;&a;&a;&a;">',
  '<!ENTITY c "&b;&b;&b;&b;&b;&b;&b;&b;&b;&b;">',
  ']><lol>&c;</lol>'))
```

Again refused at the `DOCTYPE`, so expansion never begins. There is no
expansion budget to tune, because there is no expansion.

## `allow_doctype`, and the one rule it does not relax

Real feeds do carry a bare or `PUBLIC`/`SYSTEM` doctype that declares nothing.
`allow_doctype = TRUE` accepts those and ignores the declaration:

```{r}
xml_text(xml_root(xml_parse("<!DOCTYPE feed><feed>hi</feed>",
                            allow_doctype = TRUE)))
```

An **internal subset is still rejected, even then**:

```{r, error = TRUE}
xml_parse('<!DOCTYPE d [<!ENTITY e "X">]><d>x&e;y</d>', allow_doctype = TRUE)
```

This is worth explaining, because it is a correctness rule rather than a
security one. With general entities compiled out, the parser underneath does
not record an entity declaration --- so a later reference to it is not an
error, it is passed through as *literal text*. The document above would parse
as the four characters `&e;` rather than `X`, and re-serializing would escape
it to `&amp;e;`: a different document, silently. The internal subset is the
only place a document can declare entities, so refusing it closes that hole
entirely.

An entity that was never declared anywhere is a plain parse error, which is
the correct answer for XML:

```{r, error = TRUE}
xml_parse("<p>caf&eacute;</p>")
```

If you are parsing HTML-ish content and hitting this, the answer is not a
parser option --- see the "this is XML, not HTML" section of the getting
started article.

## Limits

Everything above concerns what the document may *declare*. The limits concern
what it may *cost*. Each has its own condition class, so you can tell a hostile
document from a merely large one:

```{r, error = TRUE}
xml_parse(paste0(strrep("<a>", 300), strrep("</a>", 300)), max_depth = 256)
```

```{r, error = TRUE}
xml_parse("<a>a rather long run of text</a>", max_text = 8)
```

The full set is `max_depth`, `max_nodes`, `max_attrs`, `max_text` and
`max_memory`. The defaults are sized for documents a server might legitimately
send; lower them when you know your input is small, which turns "this parse got
expensive" into an immediate, classed error.

## Handling the errors

Every failure is an R condition under `zuxml_error`, so a caller can be as
coarse or as specific as it likes. Catching the parent class is enough to be
safe:

```{r}
parse_or_null <- function(x, ...) {
  tryCatch(xml_parse(x, ...), zuxml_error = function(e) NULL)
}
parse_or_null("<a><b></a>")
```

Distinguishing causes matters when the response is different --- a limit
suggests retrying with a bigger budget or rejecting the source outright, while
a doctype error suggests a feed you may want to allow explicitly:

```{r}
classify <- function(x, ...) {
  tryCatch({ xml_parse(x, ...); "ok" },
    zuxml_doctype_error = function(e) "doctype refused",
    zuxml_limit_error   = function(e) "too expensive",
    zuxml_parse_error   = function(e) "malformed")
}
c(classify("<a/>"),
  classify("<!DOCTYPE a><a/>"),
  classify("<a><b></a>"),
  classify("<a>xxxxxxxx</a>", max_text = 2))
```

`zuxml_limit_error` is the shared parent of the five limit conditions, so the
branch above catches all of them without naming each.

Parse errors carry the position, which is what makes a bad feed diagnosable
rather than merely rejected:

```{r}
e <- tryCatch(xml_parse("<a>\n  <b>\n</a>"), zuxml_error = identity)
conditionMessage(e)
```

## Encoding

An encoding you pass explicitly overrides whatever the document declares, which
is the behaviour an HTTP `charset` parameter should have:

```{r}
xml_text(xml_root(xml_parse(as.raw(c(0x3c, 0x61, 0x3e, 0xe9, 0x3c, 0x2f, 0x61, 0x3e)),
                            encoding = "ISO-8859-1")))
```

Encodings the parser cannot handle natively are transcoded with `iconv()`
first, and the now-stale declaration is overridden so it cannot mislead the
parser afterwards. An encoding that cannot be converted is a classed error
rather than mojibake:

```{r, error = TRUE}
xml_parse("<a>x</a>", encoding = "NOT-AN-ENCODING")
```

## What this does not protect you from

Worth being explicit, because a security vignette that only lists wins is
misleading:

- **It is not a validator.** `zuxml` checks well-formedness, not conformance to
  a schema or DTD. A document can be entirely well-formed and still be a lie
  about its contents.
- **It does not sanitize.** Text and attribute values come back as they were
  written. If you interpolate them into HTML, SQL or a shell command, that is
  your injection to prevent, not the parser's.
- **It does not fetch anything.** That is the point --- but it also means a
  document relying on an external DTD for entity definitions or attribute
  defaults will not parse the way a validating processor would. If you need
  that, you need a different parser.
- **Limits bound cost, not intent.** A document inside every limit can still be
  hostile in what it *says*.

## Summary

| Attack | Outcome |
|---|---|
| External entity (file or URL) | refused at the `DOCTYPE`; nothing opened |
| Entity amplification | refused at the `DOCTYPE`; no expansion |
| Entity declared in an internal subset | refused, even with `allow_doctype = TRUE` |
| Undeclared entity reference | `zuxml_parse_error` |
| Deep nesting, huge text, node floods | classed `zuxml_limit_error` |
| Malformed markup | `zuxml_parse_error` with line and column |

The defaults are the safe configuration. Nothing here needs to be switched on.
