---
title: "Streaming large documents"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Streaming large documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(zuxml)
```

## Who this is for

Read this if you are writing a package that receives XML in pieces --- an HTTP
response arriving in chunks, a decompressor emitting blocks --- and you would
rather not hold the whole body in memory before parsing it.

**In 0.1.0 the incremental interface is a C interface.** From R there is
`xml_parse()` for a string or raw vector and `xml_read()` for a file, and both
take the complete document. An R-level pull or callback API is planned but is
deliberately not in this release; the C seam it would sit on is what exists
today, and it is what a downstream package uses. If you are working purely in
R, this vignette will not give you anything to call --- the getting started
article covers the surface you want.

## Why the seam is C, not R

Parsing builds the tree in a C arena with no R allocation at all; R handles are
created lazily when you first touch a node. A callback-per-event API in R would
undo exactly that property, because every event would have to materialize an R
object to hand to the callback. Putting the seam in C keeps the cost where it
belongs and lets a consumer decide for itself whether an event is worth
representing.

## The contract that makes chunking safe

The guarantee worth stating first, because everything else depends on it:
**the result is byte-for-byte independent of how the input is split.** Feeding a
document as one 1 MiB buffer, as 4 KiB blocks, or one byte at a time produces
an identical tree.

That is not an aspiration. It is an acceptance criterion for the package,
tested by feeding every fixture at sizes 1, 2, 3, 7, 31 and 4096 bytes and at
random boundaries, and asserting the tree matches the whole-buffer parse ---
including splits placed inside a tag name, inside an attribute value, inside a
UTF-8 multibyte sequence, inside an entity reference and inside a CDATA
marker. One of the fuzz targets does the same with the chunk size chosen
adversarially by the fuzzer rather than enumerated.

So you may feed whatever your transport hands you. There is no buffering
you need to do first, and no minimum chunk size for correctness. (Throughput is
a different question: below about 4 KiB per call the per-call overhead starts
to show.)

## Using it from C

A consumer declares `Imports: zuxml` and `LinkingTo: zuxml`, includes the
public header, and fetches the function table once:

```c
#define ZUXML_DEFINE_API_GET
#include <zuxml.h>

static const zuxml_api *api;

void R_init_mypkg(DllInfo *dll) {
  api = zuxml_api_get();   /* aborts if the ABI version does not match */
}
```

The header exposes no Expat type, and the consumer never links against Expat.
That separation is checked on every CI run by a fixture package built exactly
the way a real consumer would be, including an assertion that the compiled
object references zero `XML_*` symbols.

To parse incrementally, create a tree builder, feed it, and finish:

```c
zux_tree_builder *b;
zux_document *doc;
zux_error err;
zux_options opt;

api->options_init(&opt);
opt.max_depth = 256;

if (api->tree_begin(&b, &opt) != ZUX_OK) { /* handle */ }

while ((n = read_some(buf, sizeof buf)) > 0) {
  if (api->tree_feed(b, buf, n) != ZUX_OK) break;   /* stop on first error */
}

if (api->tree_end(b, &doc, &err) != ZUX_OK) { /* handle */ }
```

Two things to note. Buffers passed to `tree_feed()` are caller-owned and are
never retained, so you may reuse a single stack buffer for the whole stream.
And an error is sticky: once a feed fails, the parse is over, and `tree_end()`
reports the same failure rather than a second, confusing one.

If you want events rather than a tree --- to count elements, or to pull out one
field without building anything --- `parser_feed()` is the lower seam and calls
your handlers directly.

## Errors and limits behave the same as from R

The limits in `zux_options` are enforced during feeding, not at the end, so a
hostile document fails as soon as it crosses a bound rather than after you have
paid for it. `zux_error` carries the same status, message, line and column that
the R conditions are built from, so a downstream package can map them onto its
own conditions and lose nothing:

```r
# What the R side makes of the same failures
tryCatch(xml_parse(paste0(strrep("<a>", 300), strrep("</a>", 300))),
         zuxml_error = function(e) class(e)[1:2])
```

Everything in `vignette("security")` applies unchanged to the streaming path.
It is the same parser and the same guards; only the way bytes arrive differs.
In particular the `DOCTYPE` policy is enforced on the first chunk that contains
the declaration, so a hostile document is refused early rather than after the
whole body has been read.

## A note on memory

Streaming bounds the memory used for the *input*, not for the result. The tree
is still built in full, so a 100 MiB document still produces a 100 MiB-ish
tree. What you avoid is holding the raw bytes and the tree at the same time,
which for a large response is the difference that matters.

`max_memory` bounds the arena, so a document that would build an unreasonable
tree fails with a classed error rather than exhausting the session --- and it
fails partway through feeding, not at the end.

If you want neither the bytes nor the tree, use `parser_feed()` and keep only
what you need.
