---
title: "Getting started with zuxml"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(zuxml)
```

`zuxml` parses XML into an immutable tree and gives you a small, vectorized
API to walk it. This article covers the whole public surface — there are only
about twenty functions.

## Parsing

`xml_parse()` takes a string or a raw vector; `xml_read()` takes a file path.

```{r}
doc <- xml_parse("
<catalog>
  <book id='b1'><title>The Annotated XML</title><price>29.99</price></book>
  <book id='b2'><title>Parsing for Fun</title><price>17.50</price></book>
</catalog>")

doc
```

## Navigating

Four functions cover traversal. `xml_children()` returns *every* child node,
including text and comments; `xml_elements()` returns only element children;
`xml_find()` searches descendants. All three take an optional name filter.

```{r}
root <- xml_root(doc)
books <- xml_elements(root, "book")
books

xml_find(doc, "title")      # descendants, in document order
```

Note that `xml_children()` and `xml_elements()` differ — the whitespace
between the tags above is real text content:

```{r}
xml_type(xml_children(root))
xml_type(xml_elements(root))
```

## Reading values

**Every accessor is vectorized.** Given a nodeset of length *n* you get a
result of length *n*, so there is usually no need to loop.

```{r}
xml_name(books)
xml_attr(books, "id")
xml_text(xml_elements(books, "title"))
as.numeric(xml_text(xml_elements(books, "price")))
```

`xml_attrs()` returns the full attribute set per node, as a list of named
character vectors:

```{r}
xml_attrs(books)
```

A node and a nodeset are the same object at different lengths, so the usual
vector operations work:

```{r}
length(books)
books[[1]]
xml_text(rev(books))
```

## Text and mixed content

`xml_text()` concatenates all descendant text by default. Set
`recursive = FALSE` to get only the node's own direct text children — the
distinction that matters for mixed content.

```{r}
p <- xml_root(xml_parse("<p>Hello <em>XML</em> world</p>"))

xml_text(p)
xml_text(p, recursive = FALSE)
xml_text(p, trim = TRUE)
```

## Namespaces

Names are matched on `(namespace, local name)`. Prefixes are reported but
never decide identity, so two prefixes bound to the same URI are the same
name.

```{r}
feed <- xml_parse('
<feed xmlns="http://www.w3.org/2005/Atom">
  <entry><title>First</title></entry>
  <entry><title>Second</title></entry>
</feed>')

ATOM <- "http://www.w3.org/2005/Atom"
entries <- xml_elements(xml_root(feed), "entry", ns = ATOM)
xml_text(xml_elements(entries, "title", ns = ATOM))
```

The `ns` argument has three modes:

```{r}
ns_doc <- xml_parse('<r xmlns:a="urn:a"><a:i/><i/></r>')
r <- xml_root(ns_doc)

length(xml_elements(r, "i"))                # NULL: any namespace
length(xml_elements(r, "i", ns = "urn:a"))  # that URI
length(xml_elements(r, "i", ns = NA))       # no namespace at all
```

Individual parts of a name are available separately:

```{r}
item <- xml_elements(r, "i", ns = "urn:a")
c(name = xml_name(item), local = xml_local(item),
  ns = xml_ns(item), prefix = xml_prefix(item))
```

## Writing

```{r}
xml_serialize(books[[1]])
```

Order, attributes, text and namespace *semantics* are preserved, so a
round trip is stable. Original formatting is not preserved — quote style,
whitespace between attributes and CDATA boundaries are not retained.

```{r}
once  <- xml_serialize(xml_parse("<a><b x='1'/></a>"))
twice <- xml_serialize(xml_parse(once))
identical(once, twice)
```

`xml_write()` sends the same output to a file.

## Errors are typed

Parse failures carry a position and a specific condition class, so you can
catch exactly the case you care about.

```{r, error = TRUE}
xml_parse("<a>\n  <b>\n</a>")
```

```{r}
e <- tryCatch(xml_parse("<r><a/><b/></r>", max_nodes = 2),
              zuxml_limit_error = function(e) e)
class(e)[1:3]
```

## Secure by default

`zuxml` is built for untrusted input. Document type declarations are rejected,
general-entity support is not compiled in at all, and every resource limit has
its own error class.

```{r, error = TRUE}
xml_parse('<!DOCTYPE foo [<!ENTITY xxe SYSTEM "file:///etc/passwd">]><foo>&xxe;</foo>')
```

That means external entities cannot be expanded and no file is ever opened —
the protection is structural, not a setting you have to remember. It also
means a reference to an entity the document never declared, such as `&nbsp;`,
is an error rather than silently dropped content.

`zuxml_info()` reports the policy actually compiled into your build:

```{r}
zuxml_info()
```
