---
title: "Getting started with zuxml"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(zuxml)
```

`zuxml` parses XML into an immutable tree and gives you a small, vectorized
API to walk it. This article covers the whole public surface — there are only
about twenty functions.

## This is XML, not HTML

Worth saying before you try it: **`zuxml` does not parse HTML**, and no option
changes that.

Expat is a strict, non-recovering XML parser. Real-world HTML --- `<br>`,
unquoted attributes, `&nbsp;`, unclosed `<li>` --- is not well-formed XML, so it
is an error, not something to recover from:

```{r, error = TRUE}
xml_parse("<p>a <br> b</p>")
```

Only XHTML served as well-formed XML will parse. HTML is planned for a sibling
package that reuses this one's tree and navigation API.

## Parsing

`xml_parse()` takes a string or a raw vector; `xml_read()` takes a file path.

```{r}
doc <- xml_parse("
<catalog>
  <book id='b1'><title>The Annotated XML</title><price>29.99</price></book>
  <book id='b2'><title>Parsing for Fun</title><price>17.50</price></book>
</catalog>")

doc
```

## Navigating

Five functions cover traversal. `xml_root()` gets the document element and
`xml_parent()` goes back up. `xml_children()` returns *every* child node,
including text and comments; `xml_elements()` returns only element children;
`xml_find()` searches descendants. The last two take an optional name filter —
`xml_children()` does not, because it is the "everything directly under here"
call.

```{r}
root <- xml_root(doc)
books <- xml_elements(root, "book")
books

xml_find(doc, "title")      # descendants, in document order
xml_name(xml_parent(books)) # and back up again
```

Note that `xml_children()` and `xml_elements()` differ — the whitespace
between the tags above is real text content:

```{r}
xml_type(xml_children(root))
xml_type(xml_elements(root))
```

## Reading values

**Every accessor is vectorized.** Given a nodeset of length *n* you get a
result of length *n*, so there is usually no need to loop.

```{r}
xml_name(books)
xml_attr(books, "id")
xml_text(xml_elements(books, "title"))
as.numeric(xml_text(xml_elements(books, "price")))
```

`xml_attrs()` returns the full attribute set per node, as a list of named
character vectors:

```{r}
xml_attrs(books)
```

A node and a nodeset are the same object at different lengths, so the usual
vector operations work:

```{r}
length(books)
books[[1]]
xml_text(rev(books))
```

## Text and mixed content

`xml_text()` concatenates all descendant text by default. Set
`recursive = FALSE` to get only the node's own direct text children — the
distinction that matters for mixed content.

```{r}
p <- xml_root(xml_parse("<p>Hello <em>XML</em> world</p>"))

xml_text(p)
xml_text(p, recursive = FALSE)
xml_text(p, trim = TRUE)
```

## Namespaces

Names are matched on `(namespace, local name)`. Prefixes are reported but
never decide identity, so two prefixes bound to the same URI are the same
name.

```{r}
feed <- xml_parse('
<feed xmlns="http://www.w3.org/2005/Atom">
  <entry><title>First</title></entry>
  <entry><title>Second</title></entry>
</feed>')

ATOM <- "http://www.w3.org/2005/Atom"
entries <- xml_elements(xml_root(feed), "entry", ns = ATOM)
xml_text(xml_elements(entries, "title", ns = ATOM))
```

The `ns` argument has three modes:

```{r}
ns_doc <- xml_parse('<r xmlns:a="urn:a"><a:i/><i/></r>')
r <- xml_root(ns_doc)

length(xml_elements(r, "i"))                # NULL: any namespace
length(xml_elements(r, "i", ns = "urn:a"))  # that URI
length(xml_elements(r, "i", ns = NA))       # no namespace at all
```

Individual parts of a name are available separately:

```{r}
item <- xml_elements(r, "i", ns = "urn:a")
c(name = xml_name(item), local = xml_local(item),
  ns = xml_ns(item), prefix = xml_prefix(item))
```

## Writing

```{r}
xml_serialize(books[[1]])
```

Order, attributes, text and namespace *semantics* are preserved, so a
round trip is stable. Original formatting is not preserved — quote style,
whitespace between attributes and CDATA boundaries are not retained.

```{r}
once  <- xml_serialize(xml_parse("<a><b x='1'/></a>"))
twice <- xml_serialize(xml_parse(once))
identical(once, twice)
```

`xml_write()` sends the same output to a file, and `xml_read()` reads one
back:

```{r}
f <- tempfile(fileext = ".xml")
xml_write(doc, f)
identical(xml_serialize(xml_read(f)), xml_serialize(doc))
```

## Document metadata

The XML declaration is kept, so its three fields are readable off the
document:

```{r}
decl <- xml_parse('<?xml version="1.0" encoding="UTF-8" standalone="yes"?><a/>')
c(version = xml_version(decl), encoding = xml_encoding(decl))
xml_standalone(decl)
```

## Errors are typed

Parse failures carry a position and a specific condition class, so you can
catch exactly the case you care about.

```{r, error = TRUE}
xml_parse("<a>\n  <b>\n</a>")
```

```{r}
e <- tryCatch(xml_parse("<r><a/><b/></r>", max_nodes = 2),
              zuxml_limit_error = function(e) e)
class(e)[1:3]
```

## Secure by default

`zuxml` is built for untrusted input. Document type declarations are rejected,
general-entity support is not compiled in at all, and every resource limit has
its own error class.

```{r, error = TRUE}
xml_parse('<!DOCTYPE foo [<!ENTITY xxe SYSTEM "file:///etc/passwd">]><foo>&xxe;</foo>')
```

That means external entities cannot be expanded and no file is ever opened —
the protection is structural, not a setting you have to remember. It also
means a reference to an entity the document never declared, such as `&nbsp;`,
is an error rather than silently dropped content.

`zuxml_info()` reports the policy actually compiled into your build:

```{r}
zuxml_info()
```

## For package authors

`zuxml` exposes its parser to other packages as a registered C function table,
so a downstream package can parse XML without linking against Expat itself:

```
# in DESCRIPTION
Imports:    zuxml
LinkingTo:  zuxml
```

```c
#define ZUXML_DEFINE_API_GET
#include "zuxml.h"

const zuxml_api *api = zuxml_api_get();
```

`Imports:` alone is not enough --- the consumer's `NAMESPACE` needs a real
import directive (`importFrom(zuxml, zuxml_info)`), or zuxml's namespace is
never loaded and `R_GetCCallable()` resolves nothing. The header
([`inst/include/zuxml.h`](https://github.com/pedrobtz/zuxml/blob/main/inst/include/zuxml.h))
documents the string-lifetime contract and the versioning rules; read it before
you keep any pointer a handler hands you.
